---
title: "Models"
author: "Jonathan Neimann"
date: "2024-12-04"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

```{r}
#null model 

mean_popularity <- mean(spotify30k$track_popularity, na.rm = TRUE)
spotify30k$null_prediction <- mean_popularity
mean_popularity

# Calculate residuals
spotify30k$residuals <- spotify30k$track_popularity - spotify30k$null_prediction

# Mean Squared Error (MSE)
mse <- mean(spotify30k$residuals^2, na.rm = TRUE)

# Root Mean Squared Error (RMSE)
rmse <- sqrt(mse)

# Mean Absolute Error (MAE)
mae <- mean(abs(spotify30k$residuals), na.rm = TRUE)

# Print metrics
cat("MSE:", mse, "\n")
cat("RMSE:", rmse, "\n")
cat("MAE:", mae, "\n")

null_model <- lm(track_popularity ~ 1, data = spotify30k)
summary(null_model)

```

```{r}
#simple linear model

simple_model <- lm(track_popularity ~ danceability + energy + valence, data = spotify30k)
summary(simple_model)
```
```{r}
plot(simple_model$residuals, main = "Residuals of Simple Model", ylab = "Residuals", xlab = "Index")
abline(h = 0, col = "red")
```
4. Model Fit
Residual Standard Error (RSE): 24.79

Slightly better than the null model (24.99).
R² (Multiple R-squared): 0.01615

Only ~1.6% of the variance in track_popularity is explained by this model.
This is very low, indicating a weak relationship between the predictors and the target variable.
Adjusted R²: 0.01606

Adjusted for the number of predictors; similarly low.
F-statistic: 179.1 (p-value < 2.2e-16)

The overall model is statistically significant, meaning at least one predictor has a significant relationship with track_popularity.
5. Comparison to Null Model
Null Model R²: 0 (explains no variance).
This Model R²: 0.01615 (explains ~1.6% of variance).
There’s a slight improvement, but the low R² suggests the predictors (danceability, energy, valence) alone are not strong determinants of track_popularity.

```{r}
#add genre

# Build the extended model
extended_model <- lm(track_popularity ~ danceability + energy + valence +
                     loudness + instrumentalness + factor(playlist_genre), data = spotify30k)

# View the summary
summary(extended_model)
```
```{r}
library(rstanarm)
interaction_model <- lm(track_popularity ~ danceability * energy +
                        valence * factor(playlist_genre) + loudness +
                        instrumentalness, data = spotify30k)

stan_interaction_model = stan_glm(track_popularity ~ danceability * energy +
                        valence * factor(playlist_genre) + loudness +
                        instrumentalness, data = spotify30k, family='gaussian')
# View the summary
summary(interaction_model)
summary(stan_interaction_model)
```
```{r}
anova(simple_model, extended_model, interaction_model)

# Compare AIC
AIC(simple_model, extended_model, interaction_model)
```
```{r}
plot(interaction_model$residuals, main = "Residuals of Interaction Model", ylab = "Residuals", xlab = "Index")
abline(h = 0, col = "red")
```
```{r}
predicted <- predict(interaction_model, newdata = spotify30k)

# Residuals
residuals <- spotify30k$track_popularity - predicted

# Compute metrics
mse <- mean(residuals^2, na.rm = TRUE)
rmse <- sqrt(mse)
mae <- mean(abs(residuals), na.rm = TRUE)

# Print metrics
cat("MSE:", mse, "\n")
cat("RMSE:", rmse, "\n")
cat("MAE:", mae, "\n")
```
```{r}
qqnorm(interaction_model$residuals, main = "Q-Q Plot of Residuals")
qqline(interaction_model$residuals, col = "red")  # Add reference line
hist(interaction_model$residuals, main = "Histogram of Residuals", xlab = "Residuals", col = "skyblue", breaks = 30)

pp_check(stan_interaction_model)
```


#Key influence on popularity 

```{r}
spotify30k$key <- factor(spotify30k$key)

# Model with key as a categorical variable
key_model <- lm(track_popularity ~ key, data = spotify30k)

# Summary of the model
summary(key_model)
```
Model 3: key (Categorical)
Intercept: 43.1498 → The predicted track_popularity for the reference key (Key 0, "C") is 43.15.
Coefficients:
These represent the change in track_popularity relative to the reference key ("C"):
Key 7 (-2.22): Tracks in Key 7 ("G") are associated with a ~2.22-point decrease in popularity (significant at p < 0.001).
Key 8 (+1.49): Tracks in Key 8 ("G♯/A♭") are associated with a ~1.49-point increase in popularity (significant at p < 0.05).
Other keys have coefficients that are not statistically significant, indicating no strong relationship with track_popularity relative to Key 0.
RSE: 24.98 → Almost identical to the null model.
R²: 0.001287 → This model explains only ~0.13% of the variance in track_popularity.
F-statistic: 3.833 (p = 1.532e-05) → The overall model is statistically significant, but the practical effect is minimal.
Takeaway:
While key as a categorical variable is statistically significant, its practical contribution to explaining track_popularity is negligible.
```{r}
spotify30k$key <- factor(spotify30k$key)

# Model with key and other predictors
complex_model <- lm(track_popularity ~ danceability + energy + valence + key, data = spotify30k)

# Summary of the model
summary(complex_model)
```
Significant Predictors:
danceability (7.21):
A 1-unit increase in danceability is associated with a ~7.21-point increase in track_popularity.
energy (-15.34):
A 1-unit increase in energy is associated with a ~15.34-point decrease in track_popularity.
valence (3.89):
A 1-unit increase in valence is associated with a ~3.89-point increase in track_popularity.
key7 ("G"):
Tracks in Key 7 ("G") are associated with a ~1.97-point decrease in track_popularity compared to Key 0 ("C").
key8 ("G♯/A♭"):
Tracks in Key 8 ("G♯/A♭") are associated with a ~1.72-point increase in track_popularity compared to Key 0 ("C").
3. Model Fit
Residual Standard Error (RSE): 24.78
Slight improvement over the audio-only model (24.79).
R²: 0.01726
This model explains ~1.73% of the variance in track_popularity.
Still very low, indicating a weak relationship overall.
Adjusted R²: 0.01684
Adjusted for the number of predictors; similar to R².
F-statistic: 41.05 (p < 2.2e-16)
The overall model is statistically significant.
Comparison to Previous Models
Model	R²	Adjusted R²	RSE
Null Model	0	-	24.99
Audio Features	0.01615	0.01606	24.79
Key Only	0.001287	0.0009512	24.98
Combined Model	0.01726	0.01684	24.78
The combined model provides the best performance so far, though the improvement is minimal. The key variable adds some predictive power, particularly Keys 7 and 8, but the overall explained variance remains very low.

```{r}
plot(complex_model$residuals, main = "Residuals of Combined Model", ylab = "Residuals", xlab = "Index")
abline(h = 0, col = "red")
```

#Numerical Values 
```{r}
# Identify numeric columns
numeric_vars <- spotify30k %>% select_if(is.numeric)

# Initialize an empty data frame for storing correlations
cor_results <- data.frame(var1 = character(), var2 = character(), correlation = numeric())

# Compute correlations iteratively
for (i in 1:(ncol(numeric_vars) - 1)) {
  for (j in (i + 1):ncol(numeric_vars)) {
    cor_value <- cor(numeric_vars[[i]], numeric_vars[[j]], use = "complete.obs")
    cor_results <- rbind(cor_results, 
                         data.frame(var1 = colnames(numeric_vars)[i],
                                    var2 = colnames(numeric_vars)[j],
                                    correlation = cor_value))
  }
}

# Display the top correlated pairs
cor_results <- cor_results %>% arrange(desc(abs(correlation)))
head(cor_results, 10)
```

```{r}
track_popularity_corr <- sapply(numeric_vars, function(x) cor(x, spotify30k$track_popularity, use = "complete.obs"))

# Create a data frame for easier interpretation
track_popularity_corr_df <- data.frame(
  variable = names(track_popularity_corr),
  correlation = track_popularity_corr
)

# Sort by absolute correlation values (strongest relationships first)
track_popularity_corr_df <- track_popularity_corr_df %>%
  arrange(desc(abs(correlation)))

# Display the results
print(track_popularity_corr_df)
```

```{r}
# Subset categorical variables
categorical_vars <- spotify30k %>% select_if(is.character)

# Perform chi-square test for selected variables
chi_sq <- chisq.test(table(categorical_vars$playlist_genre, categorical_vars$playlist_subgenre))

# Check results
chi_sq
```
```{r}
# Create contingency table
genre_subgenre_table <- table(spotify30k$playlist_genre, spotify30k$playlist_subgenre)

# Convert to a data frame for ggplot
heatmap_data <- as.data.frame(as.table(genre_subgenre_table))

# Plot heatmap
ggplot(heatmap_data, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Playlist Genre vs Subgenre", x = "Playlist Genre", y = "Playlist Subgenre") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
ggplot(spotify30k, aes(x = instrumentalness, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Instrumentalness (by Playlist Genre)", 
       x = "Instrumentalness", y = "Track Popularity") +
  theme_minimal()

# Scatterplot: track_popularity vs duration_ms
ggplot(spotify30k, aes(x = duration_ms, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Duration (ms) (by Playlist Genre)", 
       x = "Duration (ms)", y = "Track Popularity") +
  theme_minimal()

# Scatterplot: track_popularity vs energy
ggplot(spotify30k, aes(x = energy, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Energy (by Playlist Genre)", 
       x = "Energy", y = "Track Popularity") +
  theme_minimal()

```
```{r}
# Density plot for acousticness and track_popularity
ggplot(spotify30k, aes(x = acousticness)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  facet_wrap(~ cut(track_popularity, breaks = c(0, 25, 50, 75, 100)), scales = "free_y") +
  labs(title = "Density of Acousticness Across Popularity Levels", 
       x = "Acousticness", y = "Density")

# Density plot for instrumentalness and track_popularity
ggplot(spotify30k, aes(x = instrumentalness)) +
  geom_density(fill = "orange", alpha = 0.5) +
  facet_wrap(~ cut(track_popularity, breaks = c(0, 25, 50, 75, 100)), scales = "free_y") +
  labs(title = "Density of Instrumentalness Across Popularity Levels", 
       x = "Instrumentalness", y = "Density")
```
#Categorical Values 

```{r}
# Summary statistics: Average track_popularity by playlist_genre
genre_summary <- spotify30k %>%
  group_by(playlist_genre) %>%
  summarize(mean_popularity = mean(track_popularity, na.rm = TRUE),
            sd_popularity = sd(track_popularity, na.rm = TRUE),
            count = n()) %>%
  arrange(desc(mean_popularity))

print(genre_summary)
```
```{r}
ggplot(spotify30k, aes(x = playlist_genre, y = track_popularity, fill = playlist_genre)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(title = "Track Popularity Across Playlist Genres", 
       x = "Playlist Genre", y = "Track Popularity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(spotify30k, aes(x = playlist_subgenre, y = track_popularity, fill = playlist_subgenre)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(title = "Track Popularity Across Playlist Subgenres", 
       x = "Playlist Subgenre", y = "Track Popularity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
anova_genre <- aov(track_popularity ~ playlist_genre, data = spotify30k)
summary(anova_genre)

kruskal_genre <- kruskal.test(track_popularity ~ playlist_genre, data = spotify30k)
print(kruskal_genre)
```
ANOVA Results:
F-value (207.5):

The F-value is large, indicating substantial variability in track_popularity explained by playlist_genre.
This suggests that the genre significantly influences popularity.
P-value (< 2e-16):

The extremely low p-value shows that the differences in mean track_popularity across genres are statistically significant.
Kruskal-Wallis Test Results:
Chi-squared (1133.2):

This confirms significant differences in track_popularity rankings across genres.
P-value (< 2.2e-16):

The results are consistent with the ANOVA, confirming that genre impacts popularity even when accounting for non-normality.

#Track artist

```{r}
# ANOVA test
anova_artist <- aov(track_popularity ~ track_artist, data = spotify30k)
summary(anova_artist)
```

## mixed effects model 
```{r}
library(lme4)

# Mixed-effects model with track_artist as a random effect
model <- lmer(track_popularity ~ (1 | track_artist), data = spotify30k)

# Check variance explained by track_artist
summary(model)
```




