---
title: "Model Appendix"
author: "Jonathan Neimann"
date: "2024-12-10"
output: word_document
---
```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
spotify30k = read.csv("spotify_songs.csv")
spotify30k <- spotify30k[grepl("[a-zA-Z]", spotify30k$track_artist), ]
spotify_popularity <- spotify30k[spotify30k$track_popularity >= 5, ]
```

 # All Model Code and Results 
 
 ## Models From Report with visuals 
 
 ### Null Model
```{r}
null_model <- lm(track_popularity ~ 1, data = spotify_popularity)
summary(null_model)
plot(null_model)
```

### First Linear Model 

```{r}
lm_model <- lm(track_popularity ~ danceability + energy + valence + loudness + instrumentalness + duration_ms, 
               data = spotify_popularity)
summary(lm_model)
plot(lm_model)
```

### Interaction Linear Model 

```{r}
lm_model_interaction <- lm(track_popularity ~ 
                             energy + valence + loudness + instrumentalness + duration_ms +
                             energy:loudness + valence:energy + instrumentalness:loudness, 
                           data = spotify_popularity)
summary(lm_model_interaction)
plot(lm_model_interaction)
```

### Linear model with Genre 

```{r}
lm_model_genre <- lm(track_popularity ~ energy + valence + loudness + instrumentalness + duration_ms + danceability + factor(playlist_genre), 
                     data = spotify_popularity)
summary(lm_model_genre)
plot(lm_model_genre)
```

 ### Mixed Effects Model 
 
```{r}
library(lme4)
mixed_model <- lmer(track_popularity ~ energy + valence + loudness + instrumentalness + duration_ms + danceability + factor(playlist_genre) + 
                      (1 | track_artist), 
                    data = spotify_popularity)
summary(mixed_model)
```
 
## Additional Models 

### Logistic Regression Model 

```{r}
spotify_popularity$popularity_above_60 <- ifelse(spotify_popularity$track_popularity > 60, 1, 0)

# Fit a logistic regression model
logistic_model <- glm(popularity_above_60 ~ energy + valence + loudness + instrumentalness +
                        duration_ms + danceability + factor(playlist_genre),
                      data = spotify_popularity,
                      family = binomial)

# View the model summary
summary(logistic_model)
```
```{r}
library(ggplot2)

odds_ratios <- exp(cbind(Odds_Ratio = coef(logistic_model), confint(logistic_model)))
odds_ratios_df <- as.data.frame(odds_ratios)
odds_ratios_df$Variable <- rownames(odds_ratios_df)

ggplot(odds_ratios_df, aes(x = reorder(Variable, Odds_Ratio), y = Odds_Ratio)) +
  geom_point() +
  geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.2) +
  coord_flip() +
  labs(title = "Odds Ratios for Logistic Regression", x = "Predictor", y = "Odds Ratio") +
  theme_minimal()
```

### Linear model including "key" attribute 

```{r}
model_with_key <- lm(track_popularity ~ danceability * energy +
                                      valence * factor(playlist_genre) +
                                      instrumentalness +
                                      loudness +
                                      acousticness +
                                      duration_ms +
                                      factor(key), 
                     data = spotify_popularity)
summary(model_with_key)
```

### Mixed Model with different random effects 

```{r}
mixed_model_cross <- lmer(track_popularity ~ energy + valence + loudness + instrumentalness +
                            duration_ms + (energy | track_artist) + factor(playlist_genre), 
                          data = spotify_popularity)
summary(mixed_model_cross)
```

### Mixed Model with interaction terms 

```{r}
mixed_model_interaction <- lmer(track_popularity ~ energy * factor(playlist_genre) + valence * loudness + 
                                  instrumentalness + duration_ms + (1 | track_artist), 
                                data = spotify_popularity)
summary(mixed_model_interaction)
```

### Stan Interaction Model and pp check 

```{r}
library(rstanarm)

stan_interaction_model = stan_glm(track_popularity ~ danceability * energy +
                        valence * factor(playlist_genre) + loudness +
                        instrumentalness, data = spotify_popularity, family='gaussian', refresh = 0)

summary(stan_interaction_model)

qqnorm(stan_interaction_model$residuals, main = "Q-Q Plot of Residuals")
qqline(stan_interaction_model$residuals, col = "red")  # Add reference line
hist(stan_interaction_model$residuals, main = "Histogram of Residuals", xlab = "Residuals", col = "skyblue", breaks = 30)

pp_check(stan_interaction_model)
```

