---
title: "Vsualizations Appendix"
author: "Jonathan Neimann"
date: "2024-12-10"
output: word_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
spotify30k = read.csv("spotify_songs.csv")
spotify30k <- spotify30k[grepl("[a-zA-Z]", spotify30k$track_artist), ]
spotify_popularity <- spotify30k[spotify30k$track_popularity >= 5, ]
```

# All Visualizations and Code 

## Visualizations From Report 

### Popularity distributions 

```{r}
# Create a new dataframe where track_popularity is 5 or higher
spotify_popularity <- spotify30k[spotify30k$track_popularity >= 5, ]

hist(spotify_popularity$track_popularity, 
     main = "Distribution of Track Popularity", 
     xlab = "Track Popularity", 
     col = "blue")

```

### Coorelation Matrix 

```{r}
library(corrplot)
corr_data <- spotify_popularity %>%
  select(track_popularity, danceability, energy, loudness, valence, tempo, duration_ms, liveness, instrumentalness, tempo, speechiness, acousticness) %>%
  cor(use = "complete.obs")
corrplot(corr_data, method = "color")
```

### Popularity bar and density by Genre 

```{r}
ggplot(spotify_popularity, aes(x = playlist_genre, y = track_popularity, fill = playlist_genre)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Popularity Across Genres", x = "Genre", y = "Danceability")

ggplot(spotify_popularity, aes(x = track_popularity, fill = playlist_genre)) +
  geom_density(alpha = 0.6) +
  labs(title = "Popularity by Genre", x = "Popularity", y = "Density") +
  theme(legend.position = "bottom")
```

### Musical Keys

```{r}
key_mapping <- c("C", "C♯", "D", "D♯", "E", "F", "F♯", "G", "G♯", "A", "A♯", "B")
spotify_popularity$key_label <- key_mapping[spotify_popularity$key + 1]

key_counts <- table(spotify_popularity$key_label)

pie(key_counts, 
    main = "Proportion of Musical Keys", 
    col = rainbow(length(key_counts)))

avg_popularity_by_key <- spotify_popularity %>%
  group_by(key_label) %>%
  summarize(avg_popularity = mean(track_popularity, na.rm = TRUE))

ggplot(avg_popularity_by_key, aes(x = reorder(key_label, avg_popularity), y = avg_popularity, fill = key_label)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Average Popularity by Musical Key", x = "Key", y = "Average Popularity") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Danceability by genre

```{r}
ggplot(spotify_popularity, aes(x = danceability, fill = playlist_genre)) +
  geom_density(alpha = 0.6) +
  labs(title = "Danceability by Genre", x = "Popularity", y = "Density") +
  theme(legend.position = "bottom")
```

## Additional Visualizations 

### Playlist Genre vs Subgenre 

```{r}
# Create contingency table
genre_subgenre_table <- table(spotify_popularity$playlist_genre, spotify_popularity$playlist_subgenre)

# Convert to a data frame for ggplot
heatmap_data <- as.data.frame(as.table(genre_subgenre_table))

# Plot heatmap
ggplot(heatmap_data, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Playlist Genre vs Subgenre", x = "Playlist Genre", y = "Playlist Subgenre") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Track Popularity vs different variables scatterplots

```{r}
ggplot(spotify_popularity, aes(x = instrumentalness, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Instrumentalness (by Playlist Genre)", 
       x = "Instrumentalness", y = "Track Popularity") +
  theme_minimal()

# Scatterplot: track_popularity vs duration_ms
ggplot(spotify_popularity, aes(x = duration_ms, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Duration (ms) (by Playlist Genre)", 
       x = "Duration (ms)", y = "Track Popularity") +
  theme_minimal()

# Scatterplot: track_popularity vs energy
ggplot(spotify_popularity, aes(x = energy, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Energy (by Playlist Genre)", 
       x = "Energy", y = "Track Popularity") +
  theme_minimal()

ggplot(spotify_popularity, aes(x = danceability, y = track_popularity, color = playlist_genre)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Track Popularity vs Danceability (by Playlist Genre)", 
       x = "Danceability", y = "Track Popularity") +
  theme_minimal()
```

### Popularity by subgenre 

```{r}
ggplot(spotify_popularity, aes(x = playlist_subgenre, y = track_popularity, fill = playlist_subgenre)) +
  geom_boxplot(outlier.alpha = 0.2) +
  labs(title = "Track Popularity Across Playlist Subgenres", 
       x = "Playlist Subgenre", y = "Track Popularity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Live vs Studio 

```{r}
spotify_popularity <- spotify_popularity %>%
  mutate(recording_type = ifelse(liveness > 0.8, "Live", "Studio"))

ggplot(spotify_popularity, aes(x = recording_type, y = track_popularity, fill = recording_type)) +
  geom_boxplot() +
  labs(title = "Popularity: Live vs Studio Recordings", x = "Recording Type", y = "Popularity")
```

### popularity by tempo

```{r}
spotify_popularity <- spotify_popularity %>%
  mutate(tempo_category = cut(tempo, breaks = c(0, 90, 120, Inf), 
                              labels = c("Slow", "Medium", "Fast")))

ggplot(spotify_popularity, aes(x = tempo_category, y = track_popularity, fill = tempo_category)) +
  geom_boxplot() +
  labs(title = "Popularity by Tempo Category", x = "Tempo Category", y = "Popularity")
```

### Track Count by Tempo

```{r}
spotify_popularity <- spotify_popularity %>%
  mutate(tempo_category = cut(tempo, breaks = c(0, 90, 120, Inf), 
                              labels = c("Slow", "Medium", "Fast")))

ggplot(spotify_popularity, aes(x = tempo_category, fill = tempo_category)) +
  geom_bar() +
  labs(title = "Count of Tracks by Tempo Category", x = "Tempo Category", y = "Count")
```

### Mode Distribution 

```{r}
ggplot(spotify_popularity, aes(x = factor(mode, labels = c("Minor", "Major")), fill = factor(mode))) +
  geom_bar() +
  labs(title = "Distribution of Mode (Major vs Minor)", x = "Mode", y = "Count") +
  theme(legend.position = "none")
```

### Key DIstribution by Mode

```{r}
ggplot(spotify_popularity, aes(x = key_label, fill = factor(mode, labels = c("Minor", "Major")))) +
  geom_bar(position = "dodge") +
  labs(title = "Key Distribution by Mode", x = "Key", y = "Count", fill = "Mode") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Popularity by Key and Mode 

```{r}
avg_popularity_key_mode <- spotify_popularity %>%
  group_by(key_label, mode = factor(mode, labels = c("Minor", "Major"))) %>%
  summarize(avg_popularity = mean(track_popularity, na.rm = TRUE)) %>%
  ungroup()

ggplot(avg_popularity_key_mode, aes(x = key_label, y = mode, fill = avg_popularity)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Average Popularity by Key and Mode", x = "Key", y = "Mode", fill = "Popularity") +
  theme_minimal()
```


